\documentclass{article}
\usepackage{amsmath}
\usepackage{nicematrix}

\title{Project Euler P6}
\author{Daniel Werner}

\begin{document}

\maketitle

\section*{Problem Statement}

https://projecteuler.net/problem=6

\section*{Solution}

\begin{equation*}
    \left( \sum_{0 < i \le n} i \right)^2 
    - \sum_{0 < i \le n} i^2 =
    \sum_{0 < i, j \le n} i j
    - \sum_{0 < i = j \le n} i^2
\end{equation*}

The above on the left represents the exact formulation
from the question.  The right represents a two-index
representation of the same.  The part with
$\sum_{0 < i \le n} i^2 = \sum_{0 < i = j \le n} i^2$
is not terribly far a journey.  The part with 
$\left( \sum_{0 < i \le n} i \right)^2
= \sum_{0 < i, j \le n} i j$ is the case simply because
the latter represents a sum with any two-tuple $i$ and
$j$ from the same range 1 to n because that is what
happens when you multiply two sums of terms: each
individual term forms a square in a grid and the sum of the
entire grid represents the given two-index expression.  Here
are some small values for later comparison:

\begin{align*}
    f_0 &= 0 \\
    f_1 &= 1^2 - 1^2 = 0 \\
    f_2 &= (1 + 2)^2 - (1^2 + 2^2) = 9 - 5 = 4 \\
    f_3 &= (1 + 2 + 3)^2 - (1^2 + 2^2 + 3^2) = 36 - 14 = 22 \\
    f_4 &= (1 + 2 + 3 + 4)^2 - (1^2 + 2^2 + 3^2 + 4^2) = 100 - 30 = 70
\end{align*}

A little more
manipulation of the sum demonstrates how you can recombine
parts of these expressions:

\begin{align*}
    \sum_{0 < i, j \le n} i j
    &=
    \left(\sum_{0 < i=j \le n} i j \right)
    + 2 \cdot \left( \sum_{0 < i < j \le n} i j\right)
    \\
    &=
    \left(\sum_{0 < i = j \le n} i^2\right)
    + 2 \cdot \left(\sum_{0 < i < j \le n} i j\right)
\end{align*}

Substituting in $\left( \sum_{0 < i, j \le n} i j \right)$
in the larger expression:

\begin{align*}
    \left( \sum_{0 < i \le n} i \right)^2 
    - \left(\sum_{0 < i \le n} i^2\right) &=
    \left(\sum_{0 < i = j \le n} i^2\right)
    + 2 \left(\sum_{0 < i < j \le n} i j\right)
    - \left(\sum_{0 < i = j \le n} i^2\right)
    \\
    &= 2 \cdot \sum_{0 < i < j \le n} i j
\end{align*}

This manipulation actually turns out to be rather intuitive: the square of
the sum of the first n integers contains the sum of the first n squares,
then on top all of the other ordered two-tuples of 1 through n.  Since
there is an ij for every ji, we can just take each order-insensitive
two-tuple twice.  This sum in itself is quite simple to implement in code,
but we can go further.  Let's start by verifying some small values first though:

\begin{align*}
    f_n = 2 \cdot \sum_{0 < i < j \le n} i j
\end{align*}

\begin{align*}
    f_0 &= 2 \cdot 0 = 0 \\
    f_1 &= 2 \cdot 0 = 0 \\
    f_2 &= 2 \cdot (1 \cdot 2) = 4 \\
    f_3 &= 2 \cdot (1 \cdot 2 + 1 \cdot 3 + 2 \cdot 3) = 22 \\
    f_4 &= 2 \cdot (1 \cdot 2 + 1 \cdot 3 + 1 \cdot 4 \\
    & + 2 \cdot 3 + 2 \cdot 4 + 3 \cdot 4) = 70
\end{align*}

The small values check out nicely.  Considering the pattern setup in the checks
above shows that from step to step starting from $f_2$ the terms in the sum
remain the same, only adding new values from the last iteration, not changing
existing ones.  In $f_2$ we added $1 \cdot 2$.  In $f_3$ we added
$1 \cdot 3$ and $2 \cdot 3$.  In $f_4$ we added $1 \cdot 4$, $2 \cdot 4$,
and $3 \cdot 4$.

\begin{equation*}
    1 \cdot (n + 1) + 2 \cdot (n + 1) + ... + (n) \cdot (n + 1)
    = (n + 1) \sum_{0 < i \le n} i
\end{equation*}

Because the above captures the difference between $f_n$ and $f_{n + 1}$,
we can write a recurrence formula for $f_n$ that uses it:

\begin{align*}
    f_{n + 1} &= f_n + 2 (n + 1) \sum_{0 < i \le n} i
    \\
    &= f_n + 2 (n + 1) \frac{n^2 + n}{2}
    \\
    &= f_n + n^3 + 2n^2 + n
\end{align*}

Now this recurrence can be solved assuming the genal form (an informed guess):

\begin{equation*}
    f_n = \alpha n^4 + \beta n^3 + \gamma n^2 + \delta n + \theta
\end{equation*}

We know $f_0 = 0$ so:

\begin{align*}
    f_0 &= 0 = \alpha 0^4 + \beta 0^3 + \gamma 0^2 + \delta 0 + \theta
    \\
    \theta &= 0
\end{align*}

We know $f_1 = 0$, $f_2 = 4$, $f_3 = 22$, $f_4 = 70$, so through
 similar means:

\begin{align*}
    \alpha + \beta + \gamma + \delta = 0
    \\
    16 \alpha + 8 \beta + 4 \gamma + 2 \delta = 4
    \\
    81 \alpha + 27 \beta + 9 \gamma + 3 \delta = 22
    \\
    256 \alpha + 64 \beta + 16 \gamma + 4 \delta = 70
\end{align*}

I'm not going to lie: I used Wolfram to solve the system of equations:

\begin{align*}
    \alpha = \frac{1}{4} \\
    \beta = \frac{1}{6} \\
    \gamma = \frac{-1}{4} \\
    \delta = \frac{-1}{6}
\end{align*}

\begin{equation}
    f_n = \frac{n^4}{4} + \frac{n^3}{6} - \frac{n^2}{4} - \frac{n}{6}
    = \frac{3n^4 + 2n^3 - 3n^2 - 2n}{12} 
\end{equation}

\begin{equation*}
    f_{10} = \frac{30000 + 2000 - 300 - 20}{12} = \frac{31680}{12} = 2640
\end{equation*}

The value for $f_{10}$ checks out.

\begin{equation*}
    f_{100} = \frac{300000000 + 2000000 - 30000 - 200}{12} = 25164150
\end{equation*}


\end{document}