\documentclass{article}
\input{../shared_preamble}

\title{Project Euler - Problem 202}
\author{Solution}
\date{2025-06-23}

\begin{document}

\maketitle

\section*{Problem 202: Laserbeam}


Three mirrors are arranged in the shape of an equilateral triangle, with their reflective surfaces pointing inwards. There is an infinitesimal gap at each vertex of the triangle through which a laser beam may pass.

\par
Label the vertices $A$, $B$ and $C$. There are $2$ ways in which a laser beam may enter vertex $C$, bounce off $11$ surfaces, then exit through the same vertex: one way is shown below; the other is the reverse of that.

\par
\begin{center}
[Image: ]

\par
Note: Please refer to the original problem for this image at projecteuler.net
\end{center}
There are $80840$ ways in which a laser beam may enter vertex $C$, bounce off $1000001$ surfaces, then exit through the same vertex.

\par
In how many ways can a laser beam enter at vertex $C$, bounce off $12017639147$ surfaces, then exit through the same vertex?

\par


\section*{Notes}

\subsection*{Coordinate System Basics}

These mirror problems often rely on the idea that a 'bounce' is equivalent to continuing along a straight line while flipping the coordinates of the rest of the world.  In this model we can view the problem as involving a linear path between equivalent points on a grid.  The problem will likely turn on creating a systematic way to identify grid points and a function to count line traversals or bounces between two of them.

\par

Identify points by their displacement from the center and how many rows deep they are.  So $[0, 0]$ is the origin, $[2, 4]$ is 2 hops from the center to the right and 4 layers deep, and $[-6, 0]$ is 6 hops to the left but not any further forward than the origin, which means it cannot actually be reached.  Any vertex $[a, b]$ cannot be reached if $b < 0$ or if $|b| > a$.

\par

There is one more type of point that could be described by this coordinate system and could actually be reached in many cases, but will be excluded for the analysis.  An example of this would be $[0, 1]$, which would be halfway between A and B on the problem description.  Similarly $[1, 2]$ and $[-1, 2]$ are on lines between proper vertices.  More generally, a point $[a, b]$ is a proper vertex if $a$'s remainder when dividing by two is equal to $b$'s remainer when dividing by two.

\par

Only ${\frac{1}{3}}$ of the named vertices are valid targets -- the rest correspond with the beam hitting vertex A or vertex B after whatever number of bounces.  Drawing out the structure makes it pretty clear that if $[a, b]$ is an A vertex, then $[a, b + 2]$, $[a - 3, b + 3]$, and $[a + 3, b + 3]$ are all A analogs.  These apply recursively.  The same relation applies to B and C analogs respectively, with the required initial displacements (i.e. B starts at $[1, 1]$ and C starts at $[-1, 1]$ instead of the origin).  I would expect coordinates generated this way to comport with the requirements in the earlier paragraphs.  A few weeks later I realized as I was implementing this that it is not quite the case, as the A and B analogs occur on the left and right sides but that these functions will not reach one of those sides, depending on if A or B, starting from layer depth 1.

\section*{Implementation}
\begin{lstlisting}
(defn is-named-vertex? [[horiz vert]]
  (and (>= vert 0)
       (= (mod (abs horiz) 2) (mod (abs vert) 2))
       (<= (abs horiz) (abs vert))))
\end{lstlisting}

This function identifies valid vertices in the laserbeam's path through the grid. The conditions ensure that:
\begin{itemize}
    \item We only consider points above or on the x-axis (non-negative vertical coordinates)
    \item The coordinates maintain a consistent pattern of even/odd values as identified in the notes
    \item Points are within a triangular region where horizontal spread doesn't exceed vertical height
\end{itemize}


% TODO: Explain key parts of your implementation

\end{document}
